---
title: "Flightline Pixels Universal Kriging"
output:
  html_document:
    code_folding: hide
params:
  point_path: "GAO_point_data/gao_2016_clipped_01122016.geojson"
  boundary_path: "DMSM_vector_data/flightlines_hawaii_2016_processed.geojson"
  boundary_filter_field: "START_DATE_STR"
  boundary_filter_value: "01/12/2016"
  response_field: "Pixels"
  grid_resolution_m: 2
  seed: 52
---

This notebook fits a universal kriging model for one clipped flightline point
layer and writes a grid-based prediction surface at 2 m resolution. The
workflow validates the supplied inputs, checks coordinate trend, fits the
semivariogram first, and then performs universal kriging over the masked grid.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

```{r libraries}
suppressPackageStartupMessages({
  library(sf)
  library(sp)
  library(terra)
  library(gstat)
  library(ggplot2)
  library(knitr)
})

set.seed(params$seed)

point_path <- params$point_path
boundary_path <- params$boundary_path
boundary_filter_field <- params$boundary_filter_field
boundary_filter_value <- params$boundary_filter_value
response_field <- params$response_field
grid_resolution_m <- as.numeric(params$grid_resolution_m)
seed_value <- as.integer(params$seed)

variogram_sample_limit <- 5000L
candidate_models <- c("Sph", "Exp", "Gau", "Mat")
anisotropy_directions <- c(0, 45, 90, 135)
kriging_batch_size <- 50000L
stream_block_target_cells <- 1000000L
kriging_nmax <- 60L
minimum_point_count <- 10L
gdal_write_options <- c("COMPRESS=LZW", "TILED=YES", "BIGTIFF=YES")
```

```{r helpers}
fail_if_missing <- function(path_value, label) {
  if (!nzchar(path_value)) {
    stop(label, " must be provided via notebook parameters.", call. = FALSE)
  }
  if (!file.exists(path_value)) {
    stop("Missing ", label, ": ", path_value, call. = FALSE)
  }
}

read_valid_sf <- function(path) {
  x <- sf::st_read(path, quiet = TRUE)
  x <- sf::st_zm(x, drop = TRUE, what = "ZM")
  sf::st_make_valid(x)
}

geometry_type_set <- function(x) {
  sort(unique(as.character(sf::st_geometry_type(x, by_geometry = TRUE))))
}

require_projected_crs <- function(x, label) {
  if (is.na(sf::st_crs(x))) {
    stop(label, " must have a CRS.", call. = FALSE)
  }
  if (sf::st_is_longlat(x)) {
    stop(label, " must use a projected CRS in meters.", call. = FALSE)
  }
}

numeric_or_fail <- function(x, label) {
  y <- suppressWarnings(as.numeric(x))
  if (anyNA(y)) {
    stop(label, " must be numeric and cannot contain missing values.", call. = FALSE)
  }
  y
}

normalize_date_token <- function(x) {
  gsub("[^0-9]", "", as.character(x))
}

filter_boundary_features <- function(boundary_sf, filter_field, filter_value) {
  if (!nzchar(filter_field) || !nzchar(filter_value)) {
    return(boundary_sf)
  }

  if (!filter_field %in% names(boundary_sf)) {
    stop(
      "The boundary layer does not contain the filter field: ",
      filter_field,
      call. = FALSE
    )
  }

  boundary_values <- as.character(boundary_sf[[filter_field]])
  direct_match <- !is.na(boundary_values) & boundary_values == filter_value

  if (!any(direct_match)) {
    normalized_match <- !is.na(boundary_values) &
      normalize_date_token(boundary_values) == normalize_date_token(filter_value)
    direct_match <- normalized_match
  }

  boundary_subset <- boundary_sf[direct_match, ]

  if (nrow(boundary_subset) == 0) {
    stop(
      "No boundary features matched ",
      filter_field,
      " = ",
      filter_value,
      ".",
      call. = FALSE
    )
  }

  boundary_subset
}

extract_lm_summary <- function(model, model_label) {
  model_summary <- summary(model)
  data.frame(
    model = model_label,
    r_squared = unname(model_summary$r.squared),
    adj_r_squared = unname(model_summary$adj.r.squared),
    residual_se = unname(model_summary$sigma),
    stringsAsFactors = FALSE
  )
}

extract_lm_coefficients <- function(model, model_label) {
  coef_df <- as.data.frame(summary(model)$coefficients)
  coef_df$term <- rownames(coef_df)
  rownames(coef_df) <- NULL
  names(coef_df) <- c("estimate", "std_error", "t_value", "p_value", "term")
  coef_df$model <- model_label
  coef_df[, c("model", "term", "estimate", "std_error", "t_value", "p_value")]
}

max_pair_distance_from_sample <- function(xy) {
  if (nrow(xy) < 2) {
    stop("At least two sampled points are required to compute lag distances.", call. = FALSE)
  }

  hull_idx <- unique(chull(xy[, "X"], xy[, "Y"]))
  hull_xy <- xy[hull_idx, , drop = FALSE]

  if (nrow(hull_xy) < 2) {
    return(0)
  }

  as.numeric(max(dist(hull_xy)))
}

fit_one_variogram_model <- function(model_name, empirical_vgm, initial_psill, initial_range, initial_nugget) {
  initial_model <- if (identical(model_name, "Mat")) {
    vgm(
      psill = initial_psill,
      model = model_name,
      range = initial_range,
      nugget = initial_nugget,
      kappa = 0.5
    )
  } else {
    vgm(
      psill = initial_psill,
      model = model_name,
      range = initial_range,
      nugget = initial_nugget
    )
  }

  fit <- try(
    suppressWarnings(fit.variogram(empirical_vgm, initial_model)),
    silent = TRUE
  )

  if (inherits(fit, "try-error")) {
    return(NULL)
  }

  if (any(is.na(fit$psill)) || any(is.na(fit$range)) || any(fit$psill < 0) || any(fit$range < 0)) {
    return(NULL)
  }

  fit
}

save_base_preview <- function(pred_raster, var_raster, boundary_vect, output_path, main_prefix) {
  png(filename = output_path, width = 1800, height = 900, res = 170)
  old_par <- par(no.readonly = TRUE)
  on.exit({
    par(old_par)
    dev.off()
  }, add = TRUE)

  par(mfrow = c(1, 2), mar = c(4, 4, 4, 5))

  terra::plot(pred_raster, main = paste(main_prefix, "Prediction"))
  plot(boundary_vect, add = TRUE, border = "#202020", col = NA, lwd = 0.4)

  terra::plot(var_raster, main = paste(main_prefix, "Prediction Variance"))
  plot(boundary_vect, add = TRUE, border = "#202020", col = NA, lwd = 0.4)
}

meter_labels <- function(x) {
  paste0(format(round(x), big.mark = ","), " m")
}
```

## Parameters

```{r parameter-summary}
parameter_summary <- data.frame(
  parameter = c(
    "point_path",
    "boundary_path",
    "boundary_filter_field",
    "boundary_filter_value",
    "response_field",
    "grid_resolution_m",
    "seed"
  ),
  value = c(
    point_path,
    boundary_path,
    boundary_filter_field,
    boundary_filter_value,
    response_field,
    grid_resolution_m,
    seed_value
  ),
  stringsAsFactors = FALSE
)

kable(parameter_summary)
```

## Load And Validate Inputs

```{r load-inputs}
fail_if_missing(point_path, "point_path")
fail_if_missing(boundary_path, "boundary_path")

point_stem <- tools::file_path_sans_ext(basename(point_path))
response_slug <- tolower(gsub("[^A-Za-z0-9]+", "_", response_field))
grid_resolution_label <- gsub("\\.0+$", "", format(grid_resolution_m, trim = TRUE, scientific = FALSE))
output_dir <- file.path("outputs", paste0("flightline_uk_", point_stem))
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

points_sf <- read_valid_sf(point_path)
boundary_sf <- read_valid_sf(boundary_path)

point_geom_types <- geometry_type_set(points_sf)
boundary_geom_types <- geometry_type_set(boundary_sf)

if (!identical(point_geom_types, "POINT")) {
  stop(
    "The point layer must contain only POINT geometries; found: ",
    paste(point_geom_types, collapse = ", "),
    call. = FALSE
  )
}

if (!all(boundary_geom_types %in% c("POLYGON", "MULTIPOLYGON"))) {
  stop(
    "The boundary layer must contain only POLYGON or MULTIPOLYGON geometries; found: ",
    paste(boundary_geom_types, collapse = ", "),
    call. = FALSE
  )
}

require_projected_crs(points_sf, "The point layer")
require_projected_crs(boundary_sf, "The boundary layer")

if (sf::st_crs(boundary_sf) != sf::st_crs(points_sf)) {
  boundary_sf <- sf::st_transform(boundary_sf, sf::st_crs(points_sf))
}

boundary_sf <- filter_boundary_features(
  boundary_sf = boundary_sf,
  filter_field = boundary_filter_field,
  filter_value = boundary_filter_value
)

if (!response_field %in% names(points_sf)) {
  stop("The point layer must contain the response field: ", response_field, call. = FALSE)
}

if (!is.finite(grid_resolution_m) || grid_resolution_m <= 0) {
  stop("grid_resolution_m must be a positive numeric value.", call. = FALSE)
}

points_sf[[response_field]] <- numeric_or_fail(points_sf[[response_field]], response_field)
point_xy <- sf::st_coordinates(sf::st_geometry(points_sf))
points_sf$X <- point_xy[, "X"]
points_sf$Y <- point_xy[, "Y"]

points_sf <- points_sf[
  !sf::st_is_empty(points_sf) &
    is.finite(points_sf$X) &
    is.finite(points_sf$Y),
]

if (nrow(points_sf) < minimum_point_count) {
  stop(
    "The point layer must contain at least ",
    minimum_point_count,
    " valid points after cleaning; found ",
    nrow(points_sf),
    ".",
    call. = FALSE
  )
}

boundary_union <- sf::st_union(sf::st_geometry(boundary_sf))
boundary_union_sf <- sf::st_sf(geometry = boundary_union)

points_within_boundary <- lengths(sf::st_covered_by(points_sf, boundary_union_sf)) > 0

if (!all(points_within_boundary)) {
  stop(
    sum(!points_within_boundary),
    " supplied points are not covered by the provided boundary after CRS harmonization.",
    call. = FALSE
  )
}

analysis_summary <- data.frame(
  metric = c(
    "Point count",
    "Boundary feature count (selected)",
    "Boundary filter",
    "Response field",
    "Grid resolution",
    "CRS",
    "Boundary area (m^2)",
    "Output directory"
  ),
  value = c(
    format(nrow(points_sf), big.mark = ","),
    format(nrow(boundary_sf), big.mark = ","),
    if (nzchar(boundary_filter_field) && nzchar(boundary_filter_value)) {
      paste0(boundary_filter_field, " = ", boundary_filter_value)
    } else {
      "None"
    },
    response_field,
    paste0(grid_resolution_label, " m"),
    sf::st_crs(points_sf)$input,
    format(round(sum(as.numeric(sf::st_area(boundary_union_sf))), 2), big.mark = ","),
    output_dir
  ),
  stringsAsFactors = FALSE
)

kable(analysis_summary)
```

## Coordinate Trend Diagnostics

```{r trend-diagnostics}
trend_formula_first <- stats::as.formula(paste(response_field, "~ X + Y"))
trend_formula_second <- stats::as.formula(paste(response_field, "~ X + Y + I(X^2) + I(Y^2) + I(X * Y)"))

trend_first <- lm(trend_formula_first, data = sf::st_drop_geometry(points_sf))
trend_second <- lm(trend_formula_second, data = sf::st_drop_geometry(points_sf))

trend_model_summary <- rbind(
  extract_lm_summary(trend_first, "First-order"),
  extract_lm_summary(trend_second, "Second-order")
)

trend_coefficient_summary <- rbind(
  extract_lm_coefficients(trend_first, "First-order"),
  extract_lm_coefficients(trend_second, "Second-order")
)

kable(trend_model_summary, digits = 4)
kable(trend_coefficient_summary, digits = 4)
```

## Fit The Residual Semivariogram

```{r empirical-semivariogram}
variogram_sample_n <- min(variogram_sample_limit, nrow(points_sf))
variogram_sample_idx <- sample.int(nrow(points_sf), variogram_sample_n)
variogram_sample <- points_sf[variogram_sample_idx, ]

sample_xy <- sf::st_coordinates(sf::st_geometry(variogram_sample))
colnames(sample_xy) <- c("X", "Y")
max_sampled_pair_distance_m <- max_pair_distance_from_sample(sample_xy)
cutoff_m <- 0.4 * max_sampled_pair_distance_m
width_m <- cutoff_m / 20

if (!is.finite(cutoff_m) || cutoff_m <= 0 || !is.finite(width_m) || width_m <= 0) {
  stop("Derived semivariogram lag settings are not valid.", call. = FALSE)
}

variogram_formula <- stats::as.formula(paste(response_field, "~ X + Y"))
empirical_vgm <- variogram(
  variogram_formula,
  data = variogram_sample,
  cutoff = cutoff_m,
  width = width_m
)

if (nrow(empirical_vgm) == 0) {
  stop("The empirical semivariogram produced no lag bins.", call. = FALSE)
}

semivariogram_settings <- data.frame(
  metric = c("Variogram sample size", "Max sampled pair distance", "Cutoff", "Lag width"),
  value = c(
    format(variogram_sample_n, big.mark = ","),
    paste0(format(round(max_sampled_pair_distance_m, 2), big.mark = ","), " m"),
    paste0(format(round(cutoff_m, 2), big.mark = ","), " m"),
    paste0(format(round(width_m, 2), big.mark = ","), " m")
  ),
  stringsAsFactors = FALSE
)

kable(semivariogram_settings)
head(empirical_vgm)
```

```{r fit-semivariogram}
gamma_max <- max(empirical_vgm$gamma, na.rm = TRUE)
gamma_min <- min(empirical_vgm$gamma, na.rm = TRUE)

initial_nugget <- max(0, gamma_min * 0.5)
initial_psill <- max(.Machine$double.eps, gamma_max - initial_nugget)
initial_range <- cutoff_m / 3

fitted_models <- Filter(
  Negate(is.null),
  lapply(
    candidate_models,
    fit_one_variogram_model,
    empirical_vgm = empirical_vgm,
    initial_psill = initial_psill,
    initial_range = initial_range,
    initial_nugget = initial_nugget
  )
)

if (length(fitted_models) == 0) {
  stop("No valid semivariogram model could be fitted.", call. = FALSE)
}

fitted_sse <- sapply(fitted_models, function(fit) attr(fit, "SSErr"))
fit_labels <- sapply(fitted_models, function(fit) as.character(fit$model[2]))
names(fitted_sse) <- fit_labels

fit_summary <- data.frame(
  model = fit_labels,
  sse = as.numeric(fitted_sse),
  fitted_range_m = as.numeric(sapply(fitted_models, function(fit) fit$range[2])),
  partial_sill = as.numeric(sapply(fitted_models, function(fit) fit$psill[2])),
  nugget = as.numeric(sapply(fitted_models, function(fit) fit$psill[1])),
  stringsAsFactors = FALSE
)
fit_summary <- fit_summary[order(fit_summary$sse), ]

selected_model <- fitted_models[[which.min(fitted_sse)]]
selected_model_name <- as.character(selected_model$model[2])
fitted_range_m <- as.numeric(selected_model$range[2])

if (!is.finite(fitted_range_m) || fitted_range_m <= 0) {
  stop("The selected semivariogram model has an invalid fitted range.", call. = FALSE)
}

selected_model_summary <- data.frame(
  selected_model = selected_model_name,
  fitted_range_m = round(fitted_range_m, 2),
  maxdist_for_kriging_m = round(fitted_range_m, 2),
  nmax = kriging_nmax,
  stringsAsFactors = FALSE
)

kable(fit_summary, digits = 4)
kable(selected_model_summary, digits = 2)
selected_model
```

```{r plot-fitted-semivariogram, fig.width=9, fig.height=6}
fitted_line <- variogramLine(
  selected_model,
  maxdist = cutoff_m,
  dist_vector = seq(0, cutoff_m, by = width_m)
)

range_label <- if (fitted_range_m <= cutoff_m) {
  paste0("Fitted range = ", format(round(fitted_range_m), big.mark = ","), " m")
} else {
  paste0("Fitted range > ", format(round(cutoff_m), big.mark = ","), " m")
}

range_vline_x <- min(fitted_range_m, cutoff_m)
range_label_x <- if (fitted_range_m <= cutoff_m) {
  min(max(fitted_range_m + cutoff_m * 0.04, cutoff_m * 0.1), cutoff_m * 0.85)
} else {
  cutoff_m * 0.08
}

fit_plot <- ggplot(empirical_vgm, aes(x = dist, y = gamma)) +
  geom_point(
    aes(size = np),
    shape = 21,
    stroke = 0.8,
    color = "#5a1712",
    fill = "#912d25",
    alpha = 0.9
  ) +
  geom_line(
    data = fitted_line,
    aes(x = dist, y = gamma),
    inherit.aes = FALSE,
    color = "#1f5e5a",
    linewidth = 1.1
  ) +
  geom_vline(
    xintercept = range_vline_x,
    color = "#1f5e5a",
    linewidth = 0.9,
    linetype = "dashed"
  ) +
  annotate(
    "label",
    x = range_label_x,
    y = max(empirical_vgm$gamma, na.rm = TRUE),
    label = range_label,
    hjust = 0,
    vjust = 1,
    size = 3.8,
    color = "#173f3c",
    fill = "white",
    label.size = 0.25
  ) +
  scale_x_continuous(labels = meter_labels) +
  scale_size_continuous(name = "Pair count", range = c(1.8, 5.2)) +
  coord_cartesian(xlim = c(0, cutoff_m)) +
  labs(
    title = "Residual Semivariogram With Fitted Model",
    subtitle = paste0(
      point_stem,
      " | drift: ",
      response_field,
      " ~ X + Y | best fit: ",
      selected_model_name
    ),
    x = "Lag Distance (m)",
    y = paste0("Semivariance of ", response_field)
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "#d8e7e5", linewidth = 0.35),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#173f3c"),
    axis.text = element_text(color = "#173f3c"),
    legend.title = element_text(color = "#173f3c"),
    legend.text = element_text(color = "#173f3c"),
    plot.title = element_text(color = "#173f3c", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(color = "#315f5b", hjust = 0.5),
    plot.margin = margin(12, 12, 12, 12)
  )

fit_plot_path <- file.path(output_dir, paste0(response_slug, "_semivariogram_fit.png"))
ggsave(
  filename = fit_plot_path,
  plot = fit_plot,
  width = 9,
  height = 6,
  dpi = 300
)

fit_plot
```

## Directional Diagnostic

```{r plot-directional-semivariogram, fig.width=9, fig.height=6}
directional_vgm <- variogram(
  variogram_formula,
  data = variogram_sample,
  cutoff = cutoff_m,
  width = width_m,
  alpha = anisotropy_directions
)

directional_vgm$direction <- factor(
  paste0(directional_vgm$dir.hor, " degrees"),
  levels = paste0(anisotropy_directions, " degrees")
)

directional_plot <- ggplot(directional_vgm, aes(x = dist, y = gamma, color = direction)) +
  geom_line(linewidth = 0.8) +
  geom_point(
    shape = 21,
    size = 2.6,
    stroke = 0.7,
    fill = "white"
  ) +
  scale_color_manual(
    values = c(
      "0 degrees" = "#912d25",
      "45 degrees" = "#6f201b",
      "90 degrees" = "#b8584f",
      "135 degrees" = "#3f100d"
    ),
    name = "Direction"
  ) +
  scale_x_continuous(labels = meter_labels) +
  labs(
    title = "Directional Residual Semivariograms",
    subtitle = "Directional structure is shown for diagnosis only; the final kriging model remains isotropic.",
    x = "Lag Distance (m)",
    y = paste0("Semivariance of ", response_field)
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "#ead8d6", linewidth = 0.35),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#4d1713"),
    axis.text = element_text(color = "#4d1713"),
    legend.title = element_text(color = "#4d1713"),
    legend.text = element_text(color = "#4d1713"),
    plot.title = element_text(color = "#4d1713", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(color = "#6f201b", hjust = 0.5),
    plot.margin = margin(12, 12, 12, 12)
  )

directional_plot_path <- file.path(output_dir, paste0(response_slug, "_semivariogram_directional.png"))
ggsave(
  filename = directional_plot_path,
  plot = directional_plot,
  width = 9,
  height = 6,
  dpi = 300
)

directional_plot
```

## Build The 2 m Prediction Grid

```{r build-grid}
boundary_area_m2 <- sum(as.numeric(sf::st_area(boundary_union_sf)))
estimated_prediction_cells <- as.numeric(boundary_area_m2 / (grid_resolution_m ^ 2))

boundary_vect <- terra::vect(boundary_union_sf)
grid_template <- terra::rast(
  terra::ext(boundary_vect),
  resolution = grid_resolution_m,
  crs = sf::st_crs(points_sf)$wkt
)

mask_output_path <- file.path(output_dir, paste0("prediction_mask_", grid_resolution_label, "m.tif"))
grid_mask <- terra::rasterize(
  boundary_vect,
  grid_template,
  field = 1,
  background = NA,
  filename = mask_output_path,
  overwrite = TRUE,
  wopt = list(gdal = gdal_write_options)
)

row_block_nrows <- max(
  1L,
  as.integer(floor(stream_block_target_cells / max(1, terra::ncol(grid_mask))))
)
row_block_starts <- seq.int(1L, terra::nrow(grid_mask), by = row_block_nrows)

grid_summary <- data.frame(
  metric = c(
    "Estimated boundary cells",
    "Raster columns",
    "Raster rows",
    "Grid resolution",
    "Kriging batch size",
    "Row block size"
  ),
  value = c(
    format(round(estimated_prediction_cells), big.mark = ","),
    format(terra::ncol(grid_mask), big.mark = ","),
    format(terra::nrow(grid_mask), big.mark = ","),
    paste0(grid_resolution_label, " m"),
    format(kriging_batch_size, big.mark = ","),
    paste0(format(row_block_nrows, big.mark = ","), " rows")
  ),
  stringsAsFactors = FALSE
)

kable(grid_summary)
```

## Universal Kriging On Grid Centroids

```{r run-kriging}
points_sp <- as(points_sf[, c(response_field, "X", "Y")], "Spatial")
kriging_formula <- stats::as.formula(paste(response_field, "~ X + Y"))

pred_output_path <- file.path(output_dir, paste0(response_slug, "_uk_pred_", grid_resolution_label, "m.tif"))
var_output_path <- file.path(output_dir, paste0(response_slug, "_uk_var_", grid_resolution_label, "m.tif"))

pred_raster <- terra::rast(grid_mask)
var_raster <- terra::rast(grid_mask)

terra::writeStart(
  pred_raster,
  filename = pred_output_path,
  overwrite = TRUE,
  gdal = gdal_write_options
)
terra::writeStart(
  var_raster,
  filename = var_output_path,
  overwrite = TRUE,
  gdal = gdal_write_options
)

terra::readStart(grid_mask)

finite_prediction_cells <- 0L
finite_variance_cells <- 0L
prediction_min <- Inf
prediction_max <- -Inf
variance_min <- Inf
variance_max <- -Inf

for (row_start in row_block_starts) {
  block_nrows <- min(row_block_nrows, terra::nrow(grid_mask) - row_start + 1L)
  block_mask_values <- terra::readValues(
    grid_mask,
    row = row_start,
    nrows = block_nrows,
    col = 1,
    ncols = terra::ncol(grid_mask),
    mat = FALSE
  )

  block_prediction <- rep(NA_real_, length(block_mask_values))
  block_variance <- rep(NA_real_, length(block_mask_values))
  active_offsets <- which(!is.na(block_mask_values))

  if (length(active_offsets) > 0) {
    chunk_starts <- seq.int(1L, length(active_offsets), by = kriging_batch_size)

    for (chunk_start in chunk_starts) {
      chunk_end <- min(chunk_start + kriging_batch_size - 1L, length(active_offsets))
      chunk_offsets <- active_offsets[chunk_start:chunk_end]
      global_cells <- ((row_start - 1L) * terra::ncol(grid_mask)) + chunk_offsets
      chunk_xy <- terra::xyFromCell(grid_mask, global_cells)

      chunk_sf <- st_as_sf(
        data.frame(X = chunk_xy[, 1], Y = chunk_xy[, 2]),
        coords = c("X", "Y"),
        crs = st_crs(points_sf),
        remove = FALSE
      )
      chunk_sp <- as(chunk_sf, "Spatial")

      kriged_chunk <- krige(
        formula = kriging_formula,
        locations = points_sp,
        newdata = chunk_sp,
        model = selected_model,
        nmax = kriging_nmax,
        maxdist = fitted_range_m
      )

      block_prediction[chunk_offsets] <- kriged_chunk$var1.pred
      block_variance[chunk_offsets] <- kriged_chunk$var1.var
    }
  }

  terra::writeValues(pred_raster, block_prediction, row_start, block_nrows)
  terra::writeValues(var_raster, block_variance, row_start, block_nrows)

  finite_block_prediction <- block_prediction[is.finite(block_prediction)]
  finite_block_variance <- block_variance[is.finite(block_variance)]

  finite_prediction_cells <- finite_prediction_cells + length(finite_block_prediction)
  finite_variance_cells <- finite_variance_cells + length(finite_block_variance)

  if (length(finite_block_prediction) > 0) {
    prediction_min <- min(prediction_min, min(finite_block_prediction))
    prediction_max <- max(prediction_max, max(finite_block_prediction))
  }

  if (length(finite_block_variance) > 0) {
    variance_min <- min(variance_min, min(finite_block_variance))
    variance_max <- max(variance_max, max(finite_block_variance))
  }
}

terra::readStop(grid_mask)
terra::writeStop(pred_raster)
terra::writeStop(var_raster)
pred_raster <- terra::rast(pred_output_path)
var_raster <- terra::rast(var_output_path)

if (finite_prediction_cells == 0L) {
  stop("Universal kriging produced no finite prediction values on the grid.", call. = FALSE)
}

if (finite_variance_cells == 0L) {
  stop("Universal kriging produced no finite prediction variance values on the grid.", call. = FALSE)
}

prediction_summary <- data.frame(
  metric = c(
    "Predicted cells",
    "Cells with finite variance",
    "Prediction min",
    "Prediction max",
    "Variance min",
    "Variance max"
  ),
  value = c(
    format(finite_prediction_cells, big.mark = ","),
    format(finite_variance_cells, big.mark = ","),
    format(round(prediction_min, 4), big.mark = ","),
    format(round(prediction_max, 4), big.mark = ","),
    format(round(variance_min, 4), big.mark = ","),
    format(round(variance_max, 4), big.mark = ",")
  ),
  stringsAsFactors = FALSE
)

kable(prediction_summary)
```

## Export Preview Maps

```{r export-preview, out.width='100%'}
preview_path <- file.path(output_dir, paste0(response_slug, "_uk_preview_", grid_resolution_label, "m.png"))
save_base_preview(
  pred_raster = pred_raster,
  var_raster = var_raster,
  boundary_vect = boundary_vect,
  output_path = preview_path,
  main_prefix = paste(response_field, "UK")
)

output_summary <- data.frame(
  artifact = c(
    "Fitted semivariogram figure",
    "Directional semivariogram figure",
    "Prediction raster",
    "Variance raster",
    "Prediction preview"
  ),
  path = c(
    fit_plot_path,
    directional_plot_path,
    pred_output_path,
    var_output_path,
    preview_path
  ),
  stringsAsFactors = FALSE
)

kable(output_summary)
knitr::include_graphics(preview_path)
```
