---
title: "ROD-Only Semivariogram"
output:
  html_document:
    code_folding: hide
---

## ROD-Only Area Semivariogram

This notebook uses only the 2016 ROD center points. `Area` is the semivariogram
response, while the projected point coordinates provide meter-based lag
distances.

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(gstat)
library(ggplot2)
library(knitr)

set.seed(52)

data_path <- "GAO_point_data/2016_htmasked_brown_clipped_centers.geojson"
output_dir <- "figures"
dir.create(output_dir, showWarnings = FALSE)

rod_area_cutoff_m <- 3000
rod_area_width_m <- 180
rod_area_sample_limit <- 3000
candidate_models <- c("Sph", "Exp", "Gau", "Mat")
```

```{r load-rod-data, message=FALSE, warning=FALSE}
if (!file.exists(data_path)) {
  stop("Missing ROD data file: ", data_path)
}

rod_sf <- st_read(data_path, quiet = TRUE)

if (!"Area" %in% names(rod_sf)) {
  stop("The ROD data must contain an Area column.")
}

if (is.na(st_crs(rod_sf))) {
  stop("The ROD data must have a projected CRS so distances are in meters.")
}

if (st_is_longlat(rod_sf)) {
  stop("The ROD data uses longitude/latitude coordinates. Reproject before using meter-based lags.")
}

rod_sf$Area <- as.numeric(rod_sf$Area)
rod_coords <- st_coordinates(st_geometry(rod_sf))
rod_sf$X <- rod_coords[, "X"]
rod_sf$Y <- rod_coords[, "Y"]

rod_area_sf <- rod_sf[
  !st_is_empty(rod_sf) &
    is.finite(rod_sf$X) &
    is.finite(rod_sf$Y) &
    !is.na(rod_sf$Area),
]

stopifnot(nrow(rod_area_sf) > 1)

rod_area_sample_n <- min(rod_area_sample_limit, nrow(rod_area_sf))
rod_area_sample <- rod_area_sf[sample(nrow(rod_area_sf), rod_area_sample_n), ]

analysis_summary <- data.frame(
  metric = c(
    "ROD point count",
    "Sample count",
    "CRS",
    "Variogram cutoff",
    "Lag width"
  ),
  value = c(
    format(nrow(rod_area_sf), big.mark = ","),
    format(rod_area_sample_n, big.mark = ","),
    st_crs(rod_area_sf)$input,
    paste0(format(rod_area_cutoff_m, big.mark = ","), " m"),
    paste0(format(rod_area_width_m, big.mark = ","), " m")
  )
)

kable(analysis_summary)
```

```{r empirical-area-semivariogram, message=FALSE, warning=FALSE}
vgm_rod_area <- variogram(
  Area ~ 1,
  data = rod_area_sample,
  cutoff = rod_area_cutoff_m,
  width = rod_area_width_m
)

stopifnot(nrow(vgm_rod_area) > 0)

head(vgm_rod_area)
```

```{r fit-area-semivariogram, message=FALSE, warning=FALSE}
gamma_max <- max(vgm_rod_area$gamma, na.rm = TRUE)
gamma_min <- min(vgm_rod_area$gamma, na.rm = TRUE)

initial_nugget <- max(0, gamma_min * 0.5)
initial_psill <- max(.Machine$double.eps, gamma_max - initial_nugget)
initial_range <- rod_area_cutoff_m / 3

fit_one_model <- function(model_name) {
  initial_model <- vgm(
    psill = initial_psill,
    model = model_name,
    range = initial_range,
    nugget = initial_nugget
  )

  fit <- try(
    suppressWarnings(fit.variogram(vgm_rod_area, initial_model)),
    silent = TRUE
  )

  if (inherits(fit, "try-error") || any(is.na(fit$psill)) || any(fit$psill < 0)) {
    return(NULL)
  }

  fit
}

rod_area_fits <- Filter(
  Negate(is.null),
  lapply(candidate_models, fit_one_model)
)

stopifnot(length(rod_area_fits) > 0)

rod_area_sse <- sapply(rod_area_fits, function(fit) attr(fit, "SSErr"))
names(rod_area_sse) <- sapply(rod_area_fits, function(fit) as.character(fit$model[2]))

rod_area_fit_summary <- data.frame(
  model = names(rod_area_sse),
  sse = as.numeric(rod_area_sse)
)
rod_area_fit_summary <- rod_area_fit_summary[order(rod_area_fit_summary$sse), ]

rod_area_model <- rod_area_fits[[which.min(rod_area_sse)]]
rod_area_model_name <- as.character(rod_area_model$model[2])
rod_area_range_m <- as.numeric(rod_area_model$range[2])

selected_model_summary <- data.frame(
  selected_model = rod_area_model_name,
  fitted_range_m = round(rod_area_range_m, 1),
  cutoff_m = rod_area_cutoff_m,
  lag_width_m = rod_area_width_m
)

kable(rod_area_fit_summary)
kable(selected_model_summary)
rod_area_model
```

```{r plot-area-semivariogram, message=FALSE, warning=FALSE, fig.width=9, fig.height=6, fig.align='center'}
rod_area_fit_line <- variogramLine(
  rod_area_model,
  maxdist = rod_area_cutoff_m,
  dist_vector = seq(0, rod_area_cutoff_m, by = rod_area_width_m)
)

meter_labels <- function(x) {
  paste0(format(round(x), big.mark = ","), " m")
}

range_label <- if (rod_area_range_m <= rod_area_cutoff_m) {
  paste0("Fitted range = ", format(round(rod_area_range_m), big.mark = ","), " m")
} else {
  paste0("Fitted range > ", format(rod_area_cutoff_m, big.mark = ","), " m")
}

range_label_x <- if (rod_area_range_m <= rod_area_cutoff_m) {
  min(
    max(rod_area_range_m + rod_area_cutoff_m * 0.04, rod_area_cutoff_m * 0.12),
    rod_area_cutoff_m * 0.85
  )
} else {
  rod_area_cutoff_m * 0.08
}

range_vline_x <- min(rod_area_range_m, rod_area_cutoff_m)

rod_area_plot <- ggplot(vgm_rod_area, aes(x = dist, y = gamma)) +
  geom_point(
    aes(size = np),
    shape = 21,
    stroke = 0.8,
    color = "#5a1712",
    fill = "#912d25",
    alpha = 0.9
  ) +
  geom_line(
    data = rod_area_fit_line,
    aes(x = dist, y = gamma),
    inherit.aes = FALSE,
    color = "#1f5e5a",
    linewidth = 1.1
  ) +
  geom_vline(
    xintercept = range_vline_x,
    color = "#1f5e5a",
    linewidth = 0.9,
    linetype = "dashed"
  ) +
  annotate(
    "label",
    x = range_label_x,
    y = max(vgm_rod_area$gamma, na.rm = TRUE),
    label = range_label,
    hjust = 0,
    vjust = 1,
    size = 3.8,
    color = "#173f3c",
    fill = "white",
    label.size = 0.25
  ) +
  scale_x_continuous(labels = meter_labels) +
  scale_size_continuous(name = "Pair count", range = c(1.8, 5.2)) +
  coord_cartesian(xlim = c(0, rod_area_cutoff_m)) +
  labs(
    title = "ROD-Only Area Semivariogram",
    subtitle = paste0(
      "2016 ROD centers | best fit: ",
      rod_area_model_name,
      " | cutoff = ",
      format(rod_area_cutoff_m, big.mark = ","),
      " m, lag width = ",
      rod_area_width_m,
      " m"
    ),
    x = "Lag Distance (m)",
    y = "Semivariance of ROD Area"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "#d8e7e5", linewidth = 0.35),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#173f3c"),
    axis.text = element_text(color = "#173f3c"),
    legend.title = element_text(color = "#173f3c"),
    legend.text = element_text(color = "#173f3c"),
    plot.title = element_text(color = "#173f3c", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(color = "#315f5b", hjust = 0.5),
    plot.margin = margin(12, 12, 12, 12)
  )

ggsave(
  filename = file.path(output_dir, "rod_only_area_semivariogram_2016.png"),
  plot = rod_area_plot,
  width = 9,
  height = 6,
  dpi = 300
)

rod_area_plot
```

## X/Y Point-Distance Diagnostic

This diagnostic uses only the sampled point coordinates. It is a spacing aid for
interpreting the ROD point pattern, not a replacement for the `Area`
semivariogram above.

```{r xy-distance-diagnostic, message=FALSE, warning=FALSE, fig.width=9, fig.height=6, fig.align='center'}
sample_xy <- st_coordinates(st_geometry(rod_area_sample))
sample_distance_matrix <- as.matrix(dist(sample_xy))
diag(sample_distance_matrix) <- Inf

nearest_neighbor_m <- apply(sample_distance_matrix, 1, min)
nearest_neighbor_summary <- data.frame(
  metric = c("Minimum", "25th percentile", "Median", "75th percentile", "95th percentile", "Maximum"),
  distance_m = round(as.numeric(quantile(
    nearest_neighbor_m,
    probs = c(0, 0.25, 0.5, 0.75, 0.95, 1),
    na.rm = TRUE
  )), 1)
)

kable(nearest_neighbor_summary)

nearest_neighbor_median <- median(nearest_neighbor_m, na.rm = TRUE)

xy_distance_plot <- ggplot(
  data.frame(nearest_neighbor_m = nearest_neighbor_m),
  aes(x = nearest_neighbor_m)
) +
  geom_histogram(
    binwidth = 30,
    boundary = 0,
    fill = "#912d25",
    color = "white",
    linewidth = 0.25
  ) +
  geom_vline(
    xintercept = nearest_neighbor_median,
    color = "#1f5e5a",
    linewidth = 1,
    linetype = "dashed"
  ) +
  annotate(
    "label",
    x = nearest_neighbor_median,
    y = Inf,
    label = paste0("Median = ", format(round(nearest_neighbor_median), big.mark = ","), " m"),
    hjust = -0.05,
    vjust = 1.2,
    size = 3.8,
    color = "#173f3c",
    fill = "white",
    label.size = 0.25
  ) +
  scale_x_continuous(labels = meter_labels) +
  coord_cartesian(xlim = c(0, rod_area_cutoff_m)) +
  labs(
    title = "ROD X/Y Nearest-Neighbor Distance Diagnostic",
    subtitle = "Distances are computed from the sampled 2016 ROD point coordinates",
    x = "Nearest-Neighbor Distance (m)",
    y = "ROD Point Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "#ead8d6", linewidth = 0.35),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#4d1713"),
    axis.text = element_text(color = "#4d1713"),
    plot.title = element_text(color = "#4d1713", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(color = "#6f201b", hjust = 0.5),
    plot.margin = margin(12, 12, 12, 12)
  )

ggsave(
  filename = file.path(output_dir, "rod_only_xy_distance_diagnostic_2016.png"),
  plot = xy_distance_plot,
  width = 9,
  height = 6,
  dpi = 300
)

xy_distance_plot
```
