---
title: "Code Collapse Demonstration"
output:
  html_document:
    code_folding: hide
---

## Checking 1st and 2nd Order Trends

-   No need to remove trends from data as the $R^2$ for both $1^{st}$ and $2^{nd}$ explain less than 1% of the variance in the data

```{r trend-analysis, message=FALSE, warning=FALSE}
library(gstat)
library(sf)
library(sp)

# Load data
hawaii_sf <- st_read("./Hawaii_data/Hawaii.geojson", quiet = TRUE)
gao_sf <- st_read("./GAO_point_data/2016_htmasked_brown_clipped_centers.geojson", quiet = TRUE)

gao_sp <- as(gao_sf, "Spatial")

# Identify your target column
target <- "Area"

# Ensure it is numeric (Requirement for dissimilarity calculations)
gao_sp[[target]] <- as.numeric(gao_sp[[target]])

# Extract coordinates to X and Y columns
gao_sp$X <- coordinates(gao_sp)[,1]
gao_sp$Y <- coordinates(gao_sp)[,2]

# Check 1st and 2nd order trends
trend_check <- lm(Area ~ X + Y, data = as.data.frame(gao_sp))
summary(trend_check)

trend_check_sq <- lm(Area ~ X + Y + I(X^2) + I(Y^2) + I(X*Y), data = as.data.frame(gao_sp))
summary(trend_check_sq)

```

## Canopy Spatial Structure

```{r canopy-variogram, message=FALSE}
library(terra)

# Load Tree Canopy Cover (TCC)
tcc_2017 <- rast("./Hawaii_data/2017_tcc_hawaii.tif")

# Set zero values to NA directly on the raster
tcc_2017[tcc_2017 == 0] <- NA

# 1. Take your sample (na.rm = TRUE now safely handles the newly minted NAs)
tcc_sample <- spatSample(tcc_2017, size = 20000, method = "random", na.rm = TRUE, xy = TRUE)

# 2. RENAME the column immediately
names(tcc_sample)[3] <- "canopy"

# 3. Clean and convert
tcc_clean <- na.omit(tcc_sample)
tcc_sf <- st_as_sf(tcc_clean, coords = c("x", "y"), crs = crs(tcc_2017))

# 4. Run the variogram
vgm_tcc <- variogram(canopy ~ 1, 
                     data = tcc_sf, 
                     cutoff = 1600, 
                     width = 30)

plot(vgm_tcc, main = "Canopy Spatial Structure: Hawaii")
```

## Generating Non-ROD Pseudo-Absences (Negatives)

```{r generate-negatives, fig.width=10, fig.height=10, fig.align='center'}
# Set up the 2x2 plotting grid
par(mfrow = c(2, 2))

# ==========================================
# STEP 1: Density Map to Survey Polygon (Figure 1)
# ==========================================
density_300m <- rast("./GAO_point_data/2016_htmasked_brown_clipped_300m_countha.tif")
footprint <- density_300m * 0 + 1
survey_poly <- as.polygons(footprint) # The broad flight path

terra::plot(density_300m, main = "Step 1: 300m Density Map")

# ==========================================
# STEP 2: Clip TCC by Survey & Create Forest Polygon (Figure 2)
# ==========================================
# 1. Project boundary to match TCC (assuming tcc_2017 exists)
survey_poly_tcc <- project(survey_poly, crs(tcc_2017))

# 2. Clip TCC to flight path and apply ZERO filtering logic
tcc_clipped <- terra::crop(tcc_2017, survey_poly_tcc)
tcc_clipped <- terra::mask(tcc_clipped, survey_poly_tcc)
tcc_clipped[tcc_clipped <= 0] <- NA # <--- Keep zero filtering logic

# 3. Turn the cleaned TCC (Forest only) into a Polygon
forest_footprint <- tcc_clipped * 0 + 1
forest_poly <- as.polygons(forest_footprint)

terra::plot(tcc_clipped, main = "Step 2: 30m TCC (Filtered to Path)")

# ==========================================
# STEP 3: Sample Non-ROD Pseudo-Absences (Figure 3)
# ==========================================
# Exclude known ROD pixels before sampling pseudo-absences.
gao_sf_tcc <- sf::st_transform(gao_sf, crs = terra::crs(tcc_clipped))

if (terra::is.lonlat(tcc_clipped)) {
  stop("tcc_clipped must use a projected CRS before applying a meter-based ROD exclusion buffer.")
}

rod_exclusion_distance_m <- 30

rod_buffer <- sf::st_buffer(
  sf::st_geometry(gao_sf_tcc),
  dist = rod_exclusion_distance_m
)

rod_buffer_sf <- sf::st_sf(
  rod_mask = 1,
  geometry = rod_buffer
)

rod_mask <- terra::rasterize(
  terra::vect(rod_buffer_sf),
  tcc_clipped,
  field = "rod_mask",
  background = NA,
  touches = TRUE
)

non_rod_tcc <- tcc_clipped
non_rod_tcc[!is.na(rod_mask)] <- NA

non_rod_cell_count <- as.integer(
  terra::global(!is.na(non_rod_tcc), "sum", na.rm = TRUE)[1, 1]
)

pseudo_absence_target_n <- 265734
neg_sample_size <- min(pseudo_absence_target_n, non_rod_cell_count)
stopifnot(neg_sample_size > 0)

set.seed(52)
neg_sample <- spatSample(non_rod_tcc, size = neg_sample_size, method = "random", na.rm = TRUE, xy = TRUE)
names(neg_sample)[3] <- "canopy"
neg_sf <- st_as_sf(na.omit(neg_sample), coords = c("x", "y"), crs = crs(tcc_2017))
neg_sf$presence <- 0

neg_rod_overlap <- terra::extract(rod_mask, terra::vect(neg_sf))[, 2]
stopifnot(all(is.na(neg_rod_overlap)))

point_count <- nrow(neg_sf)
print(paste(
  "Generated",
  point_count,
  "non-ROD pseudo-absence points outside",
  rod_exclusion_distance_m,
  "m buffered known ROD areas."
))

# 2. Plot the non-ROD sampling surface as the background
terra::plot(non_rod_tcc, main = paste("Step 3:", "Non-ROD Pseudo-Absences"))

# 3. Plot non-ROD pseudo-absence points on top using high transparency 
# pch = 16 (solid circle), cex = 0.2 (very small size)
# rgb(1, 1, 1, 0.05) creates a white color that is 95% transparent
plot(st_geometry(neg_sf), 
     add = TRUE, 
     pch = 16, 
     cex = 0.2, 
     col = rgb(0.5, 0, 0.5, alpha = 0.9))
# ==========================================
# STEP 4: Plot ROD Presence Points (Figure 4)
# ==========================================
# NOTE: Replace 'rod_sf' with the actual variable name of your ROD sf object.
# ROD points already projected to match tcc_clipped above.
gao_sf_proj <- gao_sf_tcc

# 2. Plot the base raster
terra::plot(tcc_clipped, main = "Step 4: ROD Presence Points")

# 3. Add the points
# Changed pch from "." to 16 (solid circle) and added cex for size control
plot(st_geometry(gao_sf_proj), 
     add = TRUE, 
     pch = 16, 
     cex = 0.4, 
     col = "red")
```

## 30 m Positive/Negative Sampling Grid

This view zooms into a 1.6 km window so the 30 m buffer logic is visible: green cells are eligible non-ROD pseudo-absence candidates, and red cells are 30 m ROD-positive/excluded cells.

```{r sampling-grid-30m, fig.width=10, fig.height=9, fig.align='center'}
# --- 1. Choose a reproducible ROD-centered 1.6 km window ---
grid_resolution_m <- 30
rod_xy <- sf::st_coordinates(sf::st_geometry(gao_sf_proj))
zoom_center <- rod_xy[
  which.min((rod_xy[, "X"] - median(rod_xy[, "X"]))^2 +
              (rod_xy[, "Y"] - median(rod_xy[, "Y"]))^2),
  c("X", "Y")
]

zoom_radius_wide <- 800
zoom_ext_wide <- terra::ext(
  zoom_center[1] - zoom_radius_wide,
  zoom_center[1] + zoom_radius_wide,
  zoom_center[2] - zoom_radius_wide,
  zoom_center[2] + zoom_radius_wide
)

zoom_xmin <- zoom_center[1] - zoom_radius_wide
zoom_xmax <- zoom_center[1] + zoom_radius_wide
zoom_ymin <- zoom_center[2] - zoom_radius_wide
zoom_ymax <- zoom_center[2] + zoom_radius_wide

# --- 2. Build a 30 m grid for this window ---
tcc_zoom <- terra::crop(tcc_clipped, zoom_ext_wide)
tcc_30m_template <- terra::rast(
  terra::ext(tcc_zoom),
  resolution = grid_resolution_m,
  crs = terra::crs(tcc_zoom)
)
tcc_30m <- terra::resample(tcc_zoom, tcc_30m_template, method = "near")

tcc_30m <- terra::mask(tcc_30m, survey_poly_tcc)
tcc_30m[tcc_30m <= 0] <- NA

# --- 3. Rasterize the ROD exclusion mask directly onto the 30 m grid ---
zoom_bbox <- sf::st_bbox(
  c(
    xmin = zoom_xmin,
    ymin = zoom_ymin,
    xmax = zoom_xmax,
    ymax = zoom_ymax
  ),
  crs = sf::st_crs(gao_sf_proj)
)

rod_buffer_zoom_sf <- suppressWarnings(sf::st_crop(rod_buffer_sf, zoom_bbox))

if (nrow(rod_buffer_zoom_sf) > 0) {
  rod_mask_30m <- terra::rasterize(
    terra::vect(rod_buffer_zoom_sf),
    tcc_30m,
    field = "rod_mask",
    background = NA,
    touches = TRUE
  )
} else {
  rod_mask_30m <- tcc_30m * NA
}

# --- 4. Encode categorical sampling logic with stable values and colors ---
viz_logic <- tcc_30m * 0
viz_logic[!is.na(tcc_30m)] <- 1
viz_logic[!is.na(rod_mask_30m)] <- 2
names(viz_logic) <- "sampling_logic"

viz_logic <- as.factor(viz_logic)
levels(viz_logic) <- data.frame(
  ID = c(1, 2),
  Status = c("Candidate negative", "ROD positive/excluded")
)

# --- 5. Plot the categorical cells only ---
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 1), mar = c(4, 4, 4, 2))

sampling_cols <- c("#2e7d32", "#c62828")

terra::plot(
  viz_logic,
  col = sampling_cols,
  ext = zoom_ext_wide,
  legend = FALSE,
  axes = TRUE,
  main = "30 m Positive/Negative Sampling Grid"
)

par(old_par)
```

## Suspect ROD Indicator Semivariogram

```{r suspect-rod-semivariogram}
library(ggplot2)

# --- 1. Preparation ---
# Project gao_sf to match tcc_2017 CRS if needed
gao_sf <- st_transform(gao_sf, crs(tcc_2017))
pos_sf <- gao_sf
pos_sf$presence <- 1

# --- 2. Subset & Sample ---
set.seed(52)
n_samples <- min(2000, nrow(pos_sf), nrow(neg_sf))

pos_subset <- pos_sf[sample(nrow(pos_sf), n_samples), ]
neg_subset <- neg_sf[sample(nrow(neg_sf), n_samples), ]

# --- 3. Align & Merge ---
pos_final <- pos_subset[, "presence"]
neg_final <- neg_subset[, "presence"]

# Ensure column geometry names match exactly before rbind
st_geometry(neg_final) <- "geometry"

combined_sf <- rbind(pos_final, neg_final)

# Validate the merge
stopifnot(all(c(0, 1) %in% combined_sf$presence))

# --- 4. Run the Suspect ROD Indicator Semivariogram ---
rod_variogram_cutoff <- 20000
rod_variogram_width <- 180

vgm_suspect_rod <- variogram(
  presence ~ 1,
  data = combined_sf,
  cutoff = rod_variogram_cutoff,
  width = rod_variogram_width
)

ggplot(vgm_suspect_rod, aes(x = dist, y = gamma)) +
  geom_point(
    shape = 21,
    size = 3.6,
    stroke = 0.8,
    color = "#5a1712",
    fill = "#912d25"
  ) +
  labs(
    title = "Suspect ROD Indicator Semivariogram",
    x = "Lag Distance (m)",
    y = "Semivariance"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background = element_rect(fill = NA, color = NA),
    panel.background = element_rect(fill = NA, color = NA),
    panel.grid.major = element_line(color = "black", linewidth = 0.35),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#4d1713"),
    axis.text = element_text(color = "#4d1713"),
    plot.title = element_text(color = "#4d1713", face = "bold", hjust = 0.5),
    plot.margin = margin(12, 12, 12, 12)
  )
```

## ROD Area Semivariogram

This ROD-only semivariogram checks whether the mapped ROD patch area has spatial dependence among the positive observations.

```{r rod-area-semivariogram}
# --- 1. Prepare ROD-only Area Values ---
rod_area_sf <- sf::st_transform(gao_sf, crs = terra::crs(tcc_clipped))
rod_area_sf$Area <- as.numeric(rod_area_sf$Area)
rod_area_sf <- rod_area_sf[!is.na(rod_area_sf$Area), ]
stopifnot(nrow(rod_area_sf) > 0)

set.seed(52)
rod_area_sample_n <- min(2000, nrow(rod_area_sf))
rod_area_sample <- rod_area_sf[sample(nrow(rod_area_sf), rod_area_sample_n), ]

# --- 2. Run ROD Area Semivariogram ---
vgm_rod_area <- variogram(
  Area ~ 1,
  data = rod_area_sample,
  cutoff = rod_variogram_cutoff,
  width = rod_variogram_width
)

ggplot(vgm_rod_area, aes(x = dist, y = gamma)) +
  geom_point(
    shape = 21,
    size = 3.6,
    stroke = 0.8,
    color = "#5a1712",
    fill = "#912d25"
  ) +
  labs(
    title = "ROD Area Semivariogram",
    x = "Lag Distance (m)",
    y = "Semivariance"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background = element_rect(fill = NA, color = NA),
    panel.background = element_rect(fill = NA, color = NA),
    panel.grid.major = element_line(color = "black", linewidth = 0.35),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#4d1713"),
    axis.text = element_text(color = "#4d1713"),
    plot.title = element_text(color = "#4d1713", face = "bold", hjust = 0.5),
    plot.margin = margin(12, 12, 12, 12)
  )
```

## Check for Anisotropy

Directional semivariograms compare spatial structure across compass directions. Similar curves suggest isotropy, while clear separation among directions suggests anisotropy.

```{r suspect-rod-anisotropy}
# --- 1. Run Directional Semivariograms ---
anisotropy_directions <- c(0, 45, 90, 135)

vgm_suspect_rod_anisotropy <- variogram(
  presence ~ 1,
  data = combined_sf,
  cutoff = rod_variogram_cutoff,
  width = rod_variogram_width,
  alpha = anisotropy_directions
)

vgm_suspect_rod_anisotropy$direction <- factor(
  paste0(vgm_suspect_rod_anisotropy$dir.hor, " degrees"),
  levels = paste0(anisotropy_directions, " degrees")
)

# --- 2. Plot Directional Semivariograms ---
ggplot(vgm_suspect_rod_anisotropy, aes(x = dist, y = gamma, color = direction)) +
  geom_line(linewidth = 0.8) +
  geom_point(
    shape = 21,
    size = 2.8,
    stroke = 0.7,
    fill = "white"
  ) +
  scale_color_manual(
    values = c(
      "0 degrees" = "#912d25",
      "45 degrees" = "#6f201b",
      "90 degrees" = "#b8584f",
      "135 degrees" = "#3f100d"
    ),
    name = "Direction"
  ) +
  labs(
    title = "Suspect ROD Indicator Directional Semivariograms",
    x = "Lag Distance (m)",
    y = "Semivariance"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "#ead8d6", linewidth = 0.35),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#4d1713"),
    axis.text = element_text(color = "#4d1713"),
    legend.title = element_text(color = "#4d1713"),
    legend.text = element_text(color = "#4d1713"),
    plot.title = element_text(color = "#4d1713", face = "bold", hjust = 0.5),
    plot.margin = margin(12, 12, 12, 12)
  )
```
